### 2020-06-12 chapters/background.tex
Removed unecessary files
Kifer et. al - A two window paradigm algorithm for change detection
Pseudocodes for background

\begin{figure}[!htb]
    \begin{center}
      \hspace*{0.6in}
      \includegraphics[scale=0.4]{figures/sliding.png}
      \caption[True sliding window]{True sliding window, W=3, S=1}
      \label{fig:sliding-window} 
    \end{center}
\end{figure}

@@

@@

The mathematical expression below summarizes our categorization of sliding windows.
\begin{equation} 
  Window \,\, of \,\, size \,\, W, \,\, step \,\, S =
    \begin{cases}
      true \,\, sliding & \text{if \textit{S}=1}\\
      stepping & \text{if \textit{S}>1}\\
      tumbling & \text{if \textit{S}=\textit{W}}
    \end{cases}
\end{equation}

@@

Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: 
\textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Similarly, Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value.

@@

Two of the most basic SWAG algorithms are \textit{Recalculate-From-Scratch (RFS)} and \textit{Subtract-On-Evict (SOE)}.

Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm. As the name suggests, this algorithm recomputes the aggregation over the entire window for each update. For instance, the computation of a sum over a sliding window using RFS means that whenever the window changes all elements are summed. This approach works for every aggregation. However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing. Figure \ref{fig:pseudo-rfs} shows pseudocode for RFS \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.45]{figures/rfs.png}
      \caption{Recalculate-From-Scratch algorithm}
      \label{fig:pseudo-rfs}
    \end{center}
\end{figure}

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element insertion or eviction from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations. To illustrate, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Figure \ref{fig:pseudo-soe} shows pseudocode for SOE \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.45]{figures/soe.png}
      \caption{Subtract-On-Evict algorithm}
      \label{fig:pseudo-soe}
    \end{center}
\end{figure}

Subtract-On-Evict is an improvement over the Recalculate-From-Scratch algorithm that  based on the idea of aggregation invertibility. The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.

@@

Tangwongsan et al. construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfied properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

%TODO: change the sketch-like from the aggregator to the aggregation: bloom filter to membership query, CMS to frequency estimation and HLL to cardinality estimation
\begin{table}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/aggregation-operations-properties.png}
      \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
      \label{tbl:aggregations-properties}
    \end{center}
\end{table}

@@

\section{Exponential Moving Averages}
Both SOE and RFS require saving the entire window contents. IS THERE A WAY WE CAN EMULATE AN WINDOW....

>>>>

\begin{figure}[!htb]
    \begin{center}
      \hspace*{0.6in}
      \includegraphics[scale=0.35]{figures/sliding.png}
      \caption[True sliding window]{True sliding window, W=3, S=1}
      \label{fig:sliding-window} 
    \end{center}
\end{figure}

@@

\textcolor{red}{TODO change all window drawings to a not horrendous style}

@@

The mathematical expression below summarizes our categorization of sliding windows.
\begin{equation} 
\label{math:windows}
  Sliding\ window \,\, of \,\, size \,\, W, \,\, step \,\, S =
    \begin{cases}
      true \,\, sliding & \text{if \textit{S}=1}\\
      stepping & \text{if \textit{S}>1}\\
      tumbling & \text{if \textit{S}=\textit{W}}
    \end{cases}
\end{equation}

@@

Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: 
\textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Similarly, Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value. The combine function is here on forward represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.

@@

Two of the most basic SWAG algorithms are \textit{Recalculate-From-Scratch (RFS)} and \textit{Subtract-On-Evict (SOE)}. Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm. As the name suggests, this algorithm recomputes the aggregation over the entire window for each update. For instance, the computation of a sum over a sliding window using RFS means that whenever the window changes all elements are summed. This approach works for every aggregation. However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing. Recall that the combine function is represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$. Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}.

\begin{algorithm}
    \caption{Recalculate-From-Scratch insert, evict and query methods}
    \label{pseudo:rfs}
    \begin{algorithmic}[1]
        \Function{query}{vals}
            \State $agg\gets\overline{\theta}$
            \ForEach {$v \in vals$}
                \State $agg \gets agg \oplus v$
            \EndFor
            \State \Return $agg$
        \EndFunction

        \Function{insert}{vals, v}

            \State $vals.pushBack(v)$
        \EndFunction

        \Function{evict}{vals}

            \State $vals.popFront()$
        \EndFunction
    \end{algorithmic}
\end{algorithm}

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element insertion or eviction from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$. To illustrate, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Algorithm \ref{pseudo:soe} shows pseudocode for SOE.

Subtract-On-Evict is an improvement over the Recalculate-From-Scratch algorithm based on the idea of aggregation invertibility. The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.

\begin{algorithm}
    \caption{Subtract-On-Evict insert, evict and query methods}
    \label{pseudo:soe}
    \begin{algorithmic}[1]
        \Function{query}{}

            \State \Return $agg$
        \EndFunction

        \Function{insert}{vals, v}

            \State $vals.pushBack(v)$
            \State $agg \gets agg \oplus v$
        \EndFunction

        \Function{evict}{vals}

            \State $v \gets vals.popFront()$
            \State $agg \gets agg \ominus val$
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

As it turns out, besides \textit{invertibility}, there are other properties that allow us to group aggregations in different families. Tangwongsan et al. construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfied properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

\textcolor{red}{TODO: change the sketch-like from the aggregator to the aggregation: bloom filter to membership query, CMS to frequency estimation and HLL to cardinality estimation}

\begin{table}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/aggregation-operations-properties.png}
      \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
      \label{tbl:aggregations-properties}
    \end{center}
\end{table}

@@

\section{Exponential Moving Averages}

\textcolor{red}{TODO: explain EMAs, give mathematical proofs on error bounds, emulated window size based on nhalf}

### 2020-06-12 chapters/introduction.tex
Pseudocodes for background

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. Different use cases process data in different ways for different purposes. For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's\footnote{https://alibaba.com} clients, while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. This media content may have been previously recorded and stored, or may be generated on the fly. Social networks are also responsible for generating large volumes of data. A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day \cite{twitter-throughput}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. These need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the protected system.

@@

\section{Hypothesis} \label{sec:hypothesis}
%The question this Thesis aims to answer is whether sliding window aggregations and probabilistic data structures can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data with a low memory footprint.

The question this Thesis aims to answer is whether a method that relies on comparing an aggregation of the reference period and the aggregation effectively seen during the stream analysis can be used to detect data pattern shifts in the stream, in the presence of high volumes of high velocity, highly skewed and seasonal data while ensuring a low memory footprint.

\section{Hypothesis Validation} \label{sec:validation}
The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item high recall ratio --- \textit{i.e.} at least 80\%
    \item low latency --- \textit{i.e.} constant time complexity, \textit{O(1)}
    \item memory efficiency --- \textit{i.e.} constant memory complexity, \textit{O(1)}
\end{itemize}

@@

\section{Document structure} 
% TODO at the end rewrite and relink this
This section merely sums up how each of the following Chapters contributes to this Thesis.

>>>>

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. Different use cases process data in different ways for different purposes. For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's\footnote{https://alibaba.com} clients, while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. This media content may have been previously recorded and stored, or may be generated on the fly. Social networks are also responsible for generating large volumes of data. A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. These need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the protected system.

@@

\section{Hypothesis} \label{sec:hypothesis}
The question this Thesis aims to answer is whether sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.

\section{Methodology} \label{sec:methodology}

\textcolor{red}{TODO how to define success criteria without "reverse engineering" this?}

The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item high recall ratio --- \textit{i.e.} at least 80\%
    \item low latency --- \textit{i.e.} constant time complexity, \textit{O(1)}
    \item memory efficiency --- \textit{i.e.} constant memory complexity, \textit{O(1)}
\end{itemize}

@@

\section{Document structure}

\textcolor{red}{TODO at the end rewrite and relink this}

This section merely sums up how each of the following Chapters contributes to this Thesis.
